---
title: "R Introduction"
author: "István Ilyés"
date: '2019 08 28 '
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

Short R Intro
========================================================


## Introduction

### What is R?

* R is a programming language for statistical computation (developed based on the S language)
* R 1.0.0 was released in 2000, the current version is R 3.6.1 released in July 2019.

### Advantages of R
* Free and open source
* Very active user community, frequent contributions from users
  + It is easy to find a solution on the internet
* New methodologies/algorithms are available quite fastly (often developed in R first)
* It is very useful for interactive data exploring and ad hoc modeling
* Very good visualization capabilities
* It can be programmed, users are able to create customized solutions

### Drawbacks of R
* It is harder to learn it first, less user friendly than Stata/Matlab etc.
* Objects (for example data) need to be stored in physical memory 
  + It can slow down the processing and hard to handle large datasets
* No commercial support, you need to figure out yourself what went wrong

### R system
1. There is a base R system which you can download from CRAN
2. Additional packages created by users (more than 14 000)

### Resources

#### Tools related to R:
* https://cran.r-project.org/
* https://www.rstudio.com/
* https://shiny.rstudio.com/
* http://slidify.org/


#### Troubleshooting, blogs, communities:
* https://www.r-bloggers.com/
* https://stackoverflow.com/
* https://stats.stackexchange.com/


#### Books
* https://www.r-project.org/doc/bib/R-books.html
* http://www.cookbook-r.com/
* http://www.cookbook-r.com/Graphs/
* http://adv-r.had.co.nz/



#### Other tools
* https://notepad-plus-plus.org/
* https://github.com/


R training
========================================================

## R as a calculator


```{r}
3+8
```



## Objects

#### All things in R are objects 




```{r}
a <- 1
b <- 2
c <- 3

ls()  		#Printing out object names
```



```{r}
rm(a,b)  	#Remove specific objects

rm(list=ls()) #Remove all objects
```




## Data types in R

#### Vectors:
##### Several elements of a single atomic type: numeric, integer, complex, boolean, string (R does not have scalars, they are 1-element vectors)

```{r}
a <- c(1,3,6.2,6,-2,10) # numeric vector
b <- c("aaa","bbb","ccc") # character vector
c <- c(TRUE,FALSE,TRUE,FALSE,TRUE,FALSE) #logical vector

a[c(3,5)] #3rd and 5th elements of a vector, indexing
```


#### Matrices:
##### Collections of equal-length vectors

```{r}
y<-matrix(1:20, nrow=5,ncol=4)

y[,3] # 3rd column of matrix
```


```{r}
y[3,] # 3rd row of matrix 
```


```{r}
y[2:3,1:3] # rows 2,3,4 of columns 1,2,3
```



#### Lists:
##### Collections of unequal-length vectors of different types

```{r}
list1 <- list(name="Fred", mynumbers=y, mymatrix=a, age=5.3)
list2 <- list(name="David", mynumbers2=b, mymatrix2=c, age2=10)

v <- c(list1,list2) # example of a list containing two lists 
list1[[2]] # 2nd component of the list
```



```{r}
list1[["mynumbers"]] # component named mynumbers in list
```


#### Factors:
##### Categorical data (ordered, unordered)

```{r}
gender <- c(rep("male",20), rep("female", 30)) 
gender <- factor(gender) 
# stores gender as 20 1s and 30 2s and associates
# 1=female, 2=male internally (alphabetically)
# R now treats gender as a nominal variable 
summary(gender)
```



#### Data frames:
##### Data set, collections of equal-length vectors of different types

```{r}
d <- c(1,2,3,4)
e <- c("red", "white", "red", NA)
f <- c(TRUE,TRUE,TRUE,FALSE)
mydata <- data.frame(d,e,f)
names(mydata) <- c("ID","Color","Passed") # variable names

# Get columns from data frame

mydata[,2:3] # columns 2,3 of data frame

# Get rows from data frame
mydata[1:2,]
```


```{r}
mydata[c("ID","Color")] # columns ID and Age from data frame
```


```{r}
mydata$ID # variable x1 in the data frame
```


## Setting the working directory

```{r}
getwd() #which directory I am in?
```

```{r}
dir()   #what's in that directory?
```

```{r}
#Windows
setwd("C:/M/Education/Öko/Rajk kurzus/1.félév/r_intro")  #set working directory to "..."
```


## Loading and saving data

#### Built-in data:

```{r}
data() #R has built in datasets. Check them out:

data(mtcars) #loading a specific built-in dataset
```


#### Other sources:

```{r}
?read.csv
```


```{r}
?read.table
```

#### Loading Stata files:


```{r}
#install.packages("foreign") #install packages from CRAN
library(foreign) #load package in

download.file('http://fmwww.bc.edu/ec-p/data/wooldridge/affairs.dta','affairs.dta',mode='wb')

affairs_full <- read.dta('affairs.dta')
```


#### Saving data


```{r}
#CSV
write.csv(affairs_full,"affairs_full.csv")

#RData
save(affairs_full, file = "affairs.RData")

#Loading back RData
load(file = "affairs.RData")
```


Exploratory Data Analysis
=========================================


### dplyr package


* Seven basic commands: 
<b><p style="color:red">filter: filter rows by conditions</p>
<p style="color:red">slice: filter rows by index</p>
<p style="color:red">select: select columns</p>
<p style="color:red">arrange: sort rows</p>
<p style="color:red">mutate: create new variables</p>
<p style="color:red">summarise: create aggregations</p>
<p style="color:red">group_by: group by certain variables</p></b>
* Joins: inner join, left join, right join, full join 
* Window functions


### Pipe operator
* We need to use tons of parentheses in R to perform multiple operations in one command
* We can chain these operations one after another with the pipe operator:
<p style="color:red">%>%</p>




## What's in our data: basic exploring

```{r message=FALSE}
library(dplyr)
```


```{r}
#str(affairs_full)
affairs_full %>% str()
```


```{r}
#head(affairs_full,5)
affairs_full %>% head(5)
```

```{r}
#tail(affairs_full,5)
affairs_full %>% tail(5)
```


```{r}
#fix(affairs_full)
```


## Keeping, dropping columns, rows

#### R has powerful indexing features for accessing object elements. 
#### These features can be used to select and exclude variables and observations.



```{r}
#Keep columns from 1st to 9th
#affairs2 <- affairs_full[,c(1,10)]
affairs1 <- affairs_full %>% select(1:9)
affairs1 %>% head(5)
```



```{r}
#Version 1: Keep only the ID columns and the number of affairs
#affairs2 <- affairs_full[,c(1,10)] 
affairs2 <- affairs_full %>% select(1,10)
affairs2 %>% head(5)
```



```{r}
#Version2: Keep only the ID columns and the number of affairs
#affairs2 <- affairs_full[,c("id","naffairs")] 
affairs2 <- affairs_full %>% select("id","naffairs")
affairs2 %>% head(5)
```



```{r}
# Keep the first 10 rows and first 2 columns
#affairs_junk <- affairs_full[1:10,1:2] 
affairs_junk <- affairs_full %>% select(1:10) %>% slice(1:2) 

#Conditional
#affairs_junk <- affairs_full[(affairs_full$relig <= 3 & affairs_full$occup >=2),]
affairs_junk <- affairs_full %>% filter(relig<=3, occup >= 2)

# alternatively
affairs_junk <- affairs_full %>% filter(relig<=3 & occup >= 2)

# "or" condition instead of "and" condition
affairs_junk <- affairs_full %>% filter(relig<=3 | occup >= 2)
```



```{r}
# Use the "in" statement
affairs_junk <- affairs_full %>% filter(occup %in% c(4,6,7))
```


## Duplicated values


```{r}
# Check for duplicated values
#nrow(affairs_full[duplicated(affairs_full),])
affairs_full %>% duplicated() %>% sum()
```

```{r}
#affairs_dup <- rbind(affairs_full,affairs_full[1,])
affairs_dup <- affairs_full %>% slice(1) %>% rbind(affairs_full)
affairs_dup %>% duplicated() %>% sum()
```

```{r}
# Drop duplicated values
#affairs_non_dup<- affairs_dup[!duplicated(affairs_dup),]
affairs_non_dup <- affairs_dup %>% distinct(,.keep_all= TRUE)
affairs_non_dup %>% dim()
```



## Joining datasets



```{r}
# Base R commands
#append datasets
#affairs_small <- rbind(head(affairs_full,20),affairs_full[120:142,])
affairs_small <- affairs_full %>% 
  head(20) %>%
  rbind(affairs_full %>% slice(120:142))


# Simple join
#affairs_m0 <- cbind(affairs1,affairs2)
affairsm0 <- affairs1 %>% cbind(affairs2)

# Merge command
#affairs_m <- merge(affairs1,affairs2,by.x = "id", by.y = "id") 
affairs_m <- affairs1 %>% merge(affairs2,by.x = "id", by.y = "id") 
```

```{r}
# Dplyr type joins
affairs_m2 <- affairs1 %>% full_join(affairs2, by = c("id" = "id"))
affairs_m2 <- affairs1 %>% inner_join(affairs2, by = c("id" = "id"))
affairs_m2 <- affairs1 %>% left_join(affairs2, by = c("id" = "id"))
affairs_m2 <- affairs1 %>% right_join(affairs2, by = c("id" = "id"))
```

#### For futher info about dplyr joins:
* https://medium.com/@HollyEmblem/joining-data-with-dplyr-in-r-874698eb8898
* https://stat545.com/bit001_dplyr-cheatsheet.html



## Manipulating variables


### Creating new variables

```{r}
#affairs_m$ln_age=log(affairs_m$age)
affairs_m <- affairs_m %>% mutate(
  ln_age = log(age),
  age_25 = age-25
  )
```


### Creating new variables with aggregation
```{r}
#Aggregation
affairs_m <- affairs_m %>% group_by(occup) %>% mutate(mean_aff = sum(naffairs, na.rm=TRUE))
```


### Creating category variable

```{r}

affairs_m <- affairs_m %>% mutate(
  relig2 = case_when(
    relig==1 ~ 5,
    relig==2 ~ 4,
    relig==3 ~ 3,
    relig==4 ~ 2,
    relig==5 ~ 1)
)


# Using ifelse statements like excel's if function
affairs_m <- affairs_m %>% mutate(
  relig3 = ifelse(relig>=4,"Strongly religous",
                  ifelse(relig==3,"Slightly religous",
                         "Not religous"))
)

# Creating factor from continuous variable
affairs_m <- affairs_m %>% mutate(
 age2 =  cut(age,breaks=c(10,30,40,50,60))
)

affairs_m %>% as.data.frame() %>% head(10)
```



## If statement


```{r}
x <- c("what","is","truth")

if("Truth" %in% x) {
   print("Truth is found")
} else {
   print("Truth is not found")
}
```


## Creating dummy variables
#### For every unique value in the string column, create a new 1/0 column
#This is what Factors do "under-the-hood" automatically when passed to function requiring numeric data


```{r}
for(level in unique(affairs_m$occup)){
  affairs_m[paste("occup", level, sep = "_")] <- ifelse(affairs_m$occup == level, 1, 0)
}
head(affairs_m,5)
```

```{r}
library(sjmisc)

affairs_m %>% 
  to_dummy(occup,var.name="occup",suffix="label") %>%
  bind_cols(affairs_m) %>%
  select(colnames(affairs_m),everything()) %>%
  head()
```




## For loop


#### It is automating a multi-step process by organizing sequences of actions.
#### Every time some operation(s) has to be repeated, a loop may come in handy.




```{r}
x <- c("A","B","C")

for(letter in x) {
  print(letter)
}
```


## Faster loops in R

### Apply

```{r}
###Apply row-wise
apply(affairs_m["ratemarr"], 1, mean)
```


```{r}
###Apply col-wise
apply(affairs_m["ratemarr"], 2, mean)
```


```{r}
#We customize functions
apply(affairs_m["ratemarr"], 2, function(x) length(x[x<=2]))
```


### Lapply, sapply

##### we want to go through on a vector: apply a function on all points (we can do this also on list)

```{r}
#but we can get a list also with sapply
sapply(1:3, function(x) x^2, simplify=FALSE)
```


```{r}
#or we can get a vector also with lapply
lapply(1:3, function(x) x^2) %>% unlist()
```



## Other data manipulation



```{r}
#Get column names
affairs_m %>% names() #This is a character vector
```


```{r}
affairs_m %>% names() %>% data.frame() #This is a data frame, nicer format, but occupy bigger space
```


```{r}
#Get the number of columns and rows
affairs_m %>% ncol()
```


```{r}
affairs_m %>% nrow()
```

```{r}
#sorting: ascending, descending
affairs_m %>% arrange(age) %>% data.frame() %>% head()  #basic sortin
```



```{r}
affairs_m %>% arrange(desc(age)) %>% data.frame() %>% head()   #sorting decreasing
```


```{r}
#order all data frame according to var 1
affairs_m <- affairs_m %>% arrange(desc(age))       
```


```{r}
# ordering with more than 1 variables
affairs_m %>% arrange(desc(age),educ) %>% data.frame() %>% head() 
```




## Descriptive statistics


### Missing values

#### Cross tabulation, contingency tables 

```{r}
table(affairs_m$relig,affairs_m$ratemarr) 
```


```{r}
table(affairs_m$relig, useNA="ifany")  #table including missing values
```


```{r}
table(affairs_m$relig,affairs_m$ratemarr,dnn=c("Religion","Rate of marriage")) 
```


```{r}
table(affairs_m$relig,affairs_m$ratemarr,dnn=c("Religion","Rate of marriage"))/nrow(affairs_m) #relative frequency
```



```{r}
xt <- xtabs( ~ ratemarr+relig, data=affairs_m)  #Frequency by gender and admission
xt 
```


```{r}
affairs_m %>%
  group_by(male, kids) %>%
  summarize(nbr = n(),
            total = sum(naffairs, na.rm = TRUE),
            mean = mean(naffairs, na.rm = TRUE),
            median = median(naffairs, na.rm = TRUE),
            max = max(naffairs, na.rm = TRUE))
```



### Summarizing data set 


```{r}
summary(affairs_m) %>% head(5) #all variables
```


```{r}
summary(affairs_m$age) #specific variable
```

### Other statistics


```{r}
mean(affairs_m$age) #mean of age
```


```{r}
var(affairs_m$age) #variance of age
```


```{r}
quantile(affairs_m$age, na.rm=TRUE)
```


```{r}
quantile(affairs_m$age, probs=c(0.25,0.5, 0.75), na.rm=TRUE)
```


```{r}
cov(affairs_m$naffairs,affairs_m$age)
```


```{r}
cor(affairs_m$naffairs,affairs_m$age)
```


```{r}
cor.test(affairs_m$naffairs,affairs_m$age)
```



```{r}
cor.test(affairs_m$naffairs,affairs_m$age)[[1]] #get t statistics of the test
```


```{r}
cor.test(affairs_m$naffairs,affairs_m$age)[[3]] #get p value of the test
```

Visualisations 
=====================================

## Base R

### Histogram

```{r}
hist(affairs_m$age)
```



### Scatterplot

```{r}
plot(affairs_m$ratemarr,affairs_m$naffairs)
```



### Boxplot


```{r}
boxplot(affairs_m$naffairs~affairs_m$ratemarr,
        main = "Rate of marriage vs Nbr affairs",
        col="blue",
        xlab = "Rate of marriage",
        ylab = "Number of affairs")
```



### Barchart

```{r}
barplot(table(affairs_m$naffairs), main="Affair distribution")
```


### Line chart

```{r}
plot(table(affairs_m$naffairs),type="l")
```


### Several plots

```{r}
par(mfrow = c(1, 2))
plot(affairs_m$age)
plot(affairs_m$naffairs~affairs_m$age) 
```



### Saving charts


```{r results="hide"}
dev.copy(png, file = "plot1.png")
# Insert the code of the chart here
dev.off()
```




## Ggplot2 ([ggplot2](http://ggplot2.org/))

```{r warning=FALSE}
library(ggplot2)
```

### Histogram

```{r}
affairs_m %>%
  ggplot(aes(age)) +
  geom_histogram(bins=20)
```

### Scatterplot

```{r warning=FALSE}
affairs_m %>% 
  ggplot(aes(x=educ, y=age)) + 
  geom_point() 
```


```{r warning=FALSE}
affairs_m %>% 
  ggplot(aes(x=educ, y=age, color=naffairs)) + 
  geom_point(shape = 16, size = 5) +
  labs(x="Education",y="Age")+
  theme(axis.title = element_text(face="bold",size=14))
```


```{r warning=FALSE}
affairs_m %>% 
  ggplot(aes(x=educ, y=age, color=naffairs)) + 
  geom_smooth(method = "loess", se=FALSE, color="red")+
  geom_smooth(method = "lm", se=FALSE, color="forestgreen")+
  geom_point(shape = 16, size = 5) +
  labs(x="Education",y="Age")+
  theme(axis.title = element_text(face="bold",size=14))
```


### Barchart

```{r warning=FALSE}
affairs_m %>% group_by(naffairs) %>% summarize(count=n()) %>% data.frame() %>%
  ggplot(aes(x=naffairs, y=count)) + 
    geom_bar(stat="identity")
```


```{r warning=FALSE}
affairs_m %>% group_by(naffairs) %>% summarize(count=n()) %>% data.frame() %>%
  ggplot(aes(x=naffairs, y=count)) + 
    geom_bar(stat="identity") +
    coord_flip()
```

### Line chart

```{r}
affairs_m %>% group_by(naffairs) %>% summarize(count=n()) %>% data.frame() %>%
  ggplot(aes(x=naffairs, y=count)) + 
    geom_line(stat="identity")
```

### Box plot
```{r}
affairs_m %>% 
  ggplot(aes(y=naffairs, group=ratemarr)) + 
  geom_boxplot()
```

```{r warning=FALSE}
library(ggthemes)
affairs_m %>% 
  ggplot(aes(y=naffairs, group=ratemarr)) + 
  geom_boxplot(outlier.colour="red",fill="tomato1") +
  labs(title="Rate of marriage vs Nbr affairs",
       x="Rate of marriage", 
       y = "Number of affairs")+
  theme_economist()+
  theme(axis.text=element_text(size=16),
        axis.title=element_text(size=18,face="bold"),
        plot.title=element_text(size=22,face = "bold"))
```

### Visualize functions

```{r}
#Logistic function
myfun <- function(xvar) {
  1/(1 + exp(-xvar + 10))
}

#Base R
curve(myfun(x), from=0, to=20)
```

```{r}
#Ggplot2
data.frame(x=c(0, 20)) %>% 
  ggplot(aes(x=x)) + 
  stat_function(fun=myfun, geom="line")
```


#### For further info about ggplot, check out this page:
* http://r-statistics.co/Complete-Ggplot2-Tutorial-Part1-With-R-Code.html
* http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html



