---
title: "R Introduction"
author: "István Ilyés"
date: '2019 08 28 '
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

Short R Intro
========================================================


## Introduction

### What is R?

* R is a programming language for statistical computation (developed based on the S language)
* R 1.0.0 was released in 2000, the current version is R 3.6.1 released in July 2019.

### Advantages of R
* Free and open source
* Very active user community, frequent contributions from users
  + It is easy to find a solution on the internet
* New methodologies/algorithms are available quite fastly (often developed in R first)
* It is very useful for interactive data exploring and ad hoc modeling
* Very good visualization capabilities
* It can be programmed, users are able to create customized solutions

### Drawbacks of R
* It is harder to learn it first, less user friendly than Stata/Matlab etc.
* Objects (for example data) need to be stored in physical memory 
  + It can slow down the processing and hard to handle large datasets
* No commercial support, you need to figure out yourself what went wrong

### R system
1. There is a base R system which you can download from CRAN
2. Additional packages created by users (more than 14 000)

### Resources

#### Tools related to R:
* https://cran.r-project.org/
* https://www.rstudio.com/
* https://shiny.rstudio.com/
* http://slidify.org/


#### Troubleshooting, blogs, communities:
* https://www.r-bloggers.com/
* https://stackoverflow.com/
* https://stats.stackexchange.com/


#### Books
* https://www.r-project.org/doc/bib/R-books.html
* http://www.cookbook-r.com/
* http://www.cookbook-r.com/Graphs/
* http://adv-r.had.co.nz/



#### Other tools
* https://notepad-plus-plus.org/
* https://github.com/


R training
========================================================

## R as a calculator


```{r}
3+8
```



## Objects

#### All things in R are objects 




```{r}
a <- 1
b <- 2
c <- 3

ls()  		#Printing out object names
```



```{r}
rm(a,b)  	#Remove specific objects

rm(list=ls()) #Remove all objects
```




## Data types in R

#### Vectors:
##### Several elements of a single atomic type: numeric, integer, complex, boolean, string (R does not have scalars, they are 1-element vectors)

```{r}
a <- c(1,3,6.2,6,-2,10) # numeric vector
b <- c("aaa","bbb","ccc") # character vector
c <- c(TRUE,FALSE,TRUE,FALSE,TRUE,FALSE) #logical vector

a[c(3,5)] #3rd and 5th elements of a vector, indexing
```


#### Matrices:
##### Collections of equal-length vectors

```{r}
y<-matrix(1:20, nrow=5,ncol=4)

y[,3] # 3rd column of matrix
```


```{r}
y[3,] # 3rd row of matrix 
```


```{r}
y[2:3,1:3] # rows 2,3,4 of columns 1,2,3
```



#### Lists:
##### Collections of unequal-length vectors of different types

```{r}
list1 <- list(name="Fred", mynumbers=y, mymatrix=a, age=5.3)
list2 <- list(name="David", mynumbers2=b, mymatrix2=c, age2=10)

v <- c(list1,list2) # example of a list containing two lists 
list1[[2]] # 2nd component of the list
```



```{r}
list1[["mynumbers"]] # component named mynumbers in list
```


#### Factors:
##### Categorical data (ordered, unordered)

```{r}
gender <- c(rep("male",20), rep("female", 30)) 
gender <- factor(gender) 
# stores gender as 20 1s and 30 2s and associates
# 1=female, 2=male internally (alphabetically)
# R now treats gender as a nominal variable 
summary(gender)
```



#### Data frames:
##### Data set, collections of equal-length vectors of different types

```{r}
d <- c(1,2,3,4)
e <- c("red", "white", "red", NA)
f <- c(TRUE,TRUE,TRUE,FALSE)
mydata <- data.frame(d,e,f)
names(mydata) <- c("ID","Color","Passed") # variable names

mydata[,2:3] # columns 2,3,4 of data frame
```


```{r}
mydata[c("ID","Color")] # columns ID and Age from data frame
```



```{r}
mydata$ID # variable x1 in the data frame
```


## Setting the working directory

```{r}
getwd() #which directory I am in?
```

```{r}
dir()   #what's in that directory?
```

```{r}
#Windows
setwd("C:/M/Education/Öko/Rajk kurzus/1.félév/r_intro")  #set working directory to "..."
```


## Loading and saving data

#### Built-in data:

```{r}
data() #R has built in datasets. Check them out:

data(mtcars) #loading a specific built-in dataset
```


#### Other sources:

```{r}
?read.csv
```


```{r}
?read.table
```

#### Loading Stata files:


```{r}
#install.packages("foreign") #install packages from CRAN
library(foreign) #load package in

download.file('http://fmwww.bc.edu/ec-p/data/wooldridge/affairs.dta','affairs.dta',mode='wb')

affairs_full <- read.dta('affairs.dta')
```


#### Saving data


```{r}
#CSV
write.csv(affairs_full,"affairs_full.csv")

#RData
save(affairs_full, file = "affairs.RData")

#Loading back RData
load(file = "affairs.RData")
```


## What's in our data: basic exploring


```{r}
str(affairs_full)
```


```{r}
head(affairs_full,5)
```

```{r}
tail(affairs_full,5)
```


```{r}
#fix(affairs_full)
```


## Keeping, dropping columns, rows

#### R has powerful indexing features for accessing object elements. 
#### These features can be used to select and exclude variables and observations.



```{r}
affairs1 <- affairs_full[,1:9] #Keep columns from 1st to 9th
head(affairs1,5)
```



```{r}
affairs2 <- affairs_full[,c(1,10)] #Keep columns from 1st to 9th
head(affairs2,5)
```



```{r}
affairs2 <- affairs_full[,c("id","naffairs")] #Keep the id column and number of affairs
head(affairs2,5)
```


```{r}
affairs_full[,"naffairs"] #Keep number of affairs column
```



```{r}
affairs_junk <- affairs_full[1:10,1:2] #Keep the first 10 rows and first 2 columns


#Conditional
affairs_junk <- affairs_full[(affairs_full$relig <= 3 & affairs_full$occup >=2),]  #conditional subsetting with and condition
head(affairs_full[(affairs_full$relig <= 3 | affairs_full$occup >=2),],5)  #conditional subsetting with or condition
```



```{r}
head(affairs_full[which(affairs_full$relig <= 3),],5) #which function stores the indices where the statement is true
```

```{r}
head(subset(affairs_full, relig <= 3 | occup >=2, select=c(id, ratemarr)),5) #subset function is the easiest way
```


## Duplicated values


```{r}
affairs_full[duplicated(affairs_full),]
```


```{r}
#affairs_full <- affairs_full[!duplicated(affairs_full),]
```


## Joining datasets



```{r}
affairs_small <- rbind(head(affairs_full,20),affairs_full[120:142,]) #append datasets

affairs_m0 <- cbind(affairs1,affairs2) #simple join

affairs_m <- merge(affairs1,affairs2,by.x = "id", by.y = "id") 
```




## Manupulating variables


### Creating new variables

```{r}
affairs_m$ln_age=log(affairs_m$age)
```


### Creating new variables with aggregation
```{r}
#Aggregation
avg_affairs <-aggregate(affairs_m$naffairs, by=list(affairs_m$occup), FUN=mean, na.rm=TRUE)
avg_affairs
```



```{r}
sum_affairs <-aggregate(affairs_m$naffairs, by=list(affairs_m$occup), FUN=sum, na.rm=TRUE)
sum_affairs
```


```{r}
affairs_m$avg_affairs <- ave(affairs_m$naffairs, affairs_m$occup)
head(affairs_m,10)
```


#### If we want to add several type of aggregated variables to the dataset, it is recommended to use the data.table package. It also has other benefits, see [here](https://www.r-bloggers.com/intro-to-the-data-table-package/)


### Creating category variable

```{r}
#attach(affairs_m)
#detach(affairs_m)

affairs_m$relig2[affairs_m$relig==1] <- 5
affairs_m$relig2[affairs_m$relig==2] <- 4
affairs_m$relig2[affairs_m$relig==3] <- 3
affairs_m$relig2[affairs_m$relig==4] <- 2
affairs_m$relig2[affairs_m$relig==5] <- 1


# Using ifelse statements like excel's if function
affairs_m$relig3 <- ifelse(affairs_m$relig>=4,"Strongly religous",ifelse(affairs_m$relig==3,"Slightly religous","Not religous"))

# Creating factor from continuous variable
affairs_m$age2 <- cut(affairs_m$age,breaks=c(10,30,40,50,60))

head(affairs_m,10)
```



## If statement


```{r}
x <- c("what","is","truth")

if("Truth" %in% x) {
   print("Truth is found")
} else {
   print("Truth is not found")
}
```


## Creating dummy variables
#### For every unique value in the string column, create a new 1/0 column
#This is what Factors do "under-the-hood" automatically when passed to function requiring numeric data


```{r}
for(level in unique(affairs_m$occup)){
  affairs_m[paste("dummy", level, sep = "_")] <- ifelse(affairs_m$occup == level, 1, 0)
}
head(affairs_m,5)
```




## Loops


#### It is automating a multi-step process by organizing sequences of actions.
#### Every time some operation(s) has to be repeated, a loop may come in handy.




```{r}
x <- c("A","B","C")

for(letter in x) {
  print(letter)
}
```


## Faster loops in R

### Apply

```{r}
###Apply row-wise
apply(affairs_m["ratemarr"], 1, mean)
```


```{r}
###Apply col-wise
apply(affairs_m["ratemarr"], 2, mean)
```


```{r}
#We customize functions
apply(affairs_m["ratemarr"], 2, function(x) length(x[x<=2]))
```


### Lapply, sapply

##### we want to go through on a vector: apply a function on all points (we can do this also on list)


```{r}
#sapply will return a vector
head(sapply(affairs_m["ratemarr"], function(x) x^2),5)
```



```{r}
affairs_m$ratemarr_sq1 <- sapply(affairs_m["ratemarr"], function(x) x^2)

#lapply will return a list
head(lapply(affairs_m["ratemarr"], function(x) x^2),5)
```


```{r}
affairs_m$ratemarr_sq2 <- lapply(affairs_m["ratemarr"], function(x) x^2)


#but we can get a list also with sapply
sapply(1:3, function(x) x^2, simplify=FALSE)
```


```{r}
#or we can get a vector also with lapply
unlist(lapply(1:3, function(x) x^2))
```



## Other data manipulation



```{r}
#Get column names
names(affairs_m) #This is a character vector
```


```{r}
data.frame("Names"=names(affairs_m)) #This is a data frame, nicer format, but occupy bigger space
```


```{r}
#Get the number of columns and rows
ncol(affairs_m)
```


```{r}
nrow(affairs_m)
```

```{r}
#sorting: ascending, descending
head(sort(affairs_m$age),5)                     #basic sortin
```



```{r}
head(sort(affairs_m$age, decreasing=TRUE),5)    #sorting decreasing
```


```{r}
head(sort(affairs_m$age, na.last=TRUE),5)       #put missing values to the end
```


```{r}
head(affairs_m[order(affairs_m$age),1:(ncol(affairs_m)-1)],5)        #order all data frame according to var 1
```


```{r}
head(affairs_m[order(affairs_m$age, affairs_m$relig),1:(ncol(affairs_m)-1)],5)    #ordering with more variables
```




## Descriptive statistics


### Missing values

#### Cross tabulation, contingency tables 

```{r}
table(affairs_m$relig,affairs_m$ratemarr) 
```


```{r}
table(affairs_m$relig, useNA="ifany")  #table including missing values
```


```{r}
table(affairs_m$relig,affairs_m$ratemarr,dnn=c("Religion","Rate of marriage")) 
```


```{r}
table(affairs_m$relig,affairs_m$ratemarr,dnn=c("Religion","Rate of marriage"))/nrow(affairs_m) #relative frequency
```



```{r}
xt <- xtabs( ~ ratemarr+relig, data=affairs_m)  #Frequency by gender and admission
xt 
```






### Summarizing data set 


```{r}
head(summary(affairs_m),5) #all variables
```


```{r}
summary(affairs_m$age) #specific variable
```

### Other statistics


```{r}
mean(affairs_m$age) #mean of age
```


```{r}
var(affairs_m$age) #variance of age
```


```{r}
quantile(affairs_m$age, na.rm=TRUE)
```


```{r}
quantile(affairs_m$age, probs=c(0.25,0.5, 0.75), na.rm=TRUE)
```


```{r}
cov(affairs_m$naffairs,affairs_m$age)
```


```{r}
cor(affairs_m$naffairs,affairs_m$age)
```


```{r}
cor.test(affairs_m$naffairs,affairs_m$age)
```



```{r}
cor.test(affairs_m$naffairs,affairs_m$age)[[1]] #get t statistics of the test
```


```{r}
cor.test(affairs_m$naffairs,affairs_m$age)[[3]] #get p value of the test
```

Visualisations 
=====================================

## Base R

### Histogram

```{r}
hist(affairs_m$age)
```

```{r}
with(affairs_m, hist(age))
```



### Scatterplot

```{r}
plot(affairs_m$yrsmarr,affairs_m$naffairs)
```

```{r}
plot(affairs_m$age,affairs_m$naffairs)
```

```{r}
plot(affairs_m$relig,affairs_m$naffairs)
```

```{r}
plot(affairs_m$ratemarr,affairs_m$naffairs)
```



### Boxplot


```{r}
boxplot(affairs_m$naffairs~affairs_m$ratemarr,
        main = "Rate of marriage vs Nbr affairs",
        col="blue",
        xlab = "Rate of marriage",
        ylab = "Number of affairs")
```



### Barchart

```{r}
barplot(table(affairs_m$naffairs), main="Affair distribution")
```


### Line chart

```{r}
plot(table(affairs_m$naffairs),type="l")
```


### Several plots


```{r}
par(mfrow = c(1, 2))
plot(affairs_m$age)
plot(affairs_m$naffairs~affairs_m$age) 
```



### Saving charts


```{r}
par(mfrow = c(1, 1))

boxplot(affairs_m$naffairs~affairs_m$ratemarr,
        main = "Rate of marriage vs Nbr affairs",
        col="blue",
        xlab = "Rate of marriage",
        ylab = "Number of affairs")
```

```{r}
dev.copy(png, file = "plot1.png")
```



```{r}
dev.off()
```



## Ggplot2 ([ggplot2](http://ggplot2.org/))

```{r}
library(ggplot2)
```


```{r}
p <- ggplot(mtcars, aes(x=wt, y=mpg)) + geom_point()
p
```


```{r}
# Lines and points together
ggplot(pressure, aes(x=temperature, y=pressure)) + geom_line() + geom_point()
```


```{r}
ggplot(BOD, aes(x=Time, y=demand)) + geom_bar(stat="identity")
```

```{r}
ggplot(ToothGrowth, aes(x=supp, y=len)) + geom_boxplot()
```


```{r}
#Logistic function
myfun <- function(xvar) {
  1/(1 + exp(-xvar + 10))
}

#Base R
curve(myfun(x), from=0, to=20)
```

```{r}
#Ggplot2
ggplot(data.frame(x=c(0, 20)), aes(x=x)) + stat_function(fun=myfun, geom="line")
```






